---
title: "Class 11:Comparative Structure Analysis"
author: "Emily Chen (PID: A16925878)"
format: pdf
---

Load up the packages we will need for the analysis of protein structure sets.

```{r}
library(bio3d)
```

We will analyze the ADK starting with a single ADK database accession code: "1ake_A"

```{r}
# id <- get.seq("1ake_A")
id <- '1ake_A'
aa <- get.seq(id)
aa
```

Now we can search the PDB database to find all related entries
```{r}
blast <-blast.pdb(aa)
```

```{r}
head(blast$hit.tbl)
```

make a little summary figure of these results

```{r}
hits <- plot(blast)
```

The "top hits" are our most similar entries in the resultst

```{r}
hits$pdb.id
```

```{r}
#download related PDB files
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```

```{r}
# Align related PBDs
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```

```{r}
library(bio3dview)
#view.pdbs(pdbs)
```

This view is better than all the data; however, it is still hard to see the alignment and to see which one is more similar to the other or how to families of sequences are similar.

Lets do a PCA:
```{r}
pc<- pca(pdbs)
plot(pc)
```

```{r}
plot(pc,pc.axes=1:2)
```




Write a PDB "trajectory" for mol-star
```{r}
mktrj(pc, fil="pca_results.pdb")
```



